---
title: "Peer Assessment 2"
output: html_document
---

# Title: 
 Natural disasters in the USA: Their impact on People and goods


## synopsis

This document provides a simple analysis of weather events and their negative impacts on both the population and the economy in the United States. The analysed data are provided by the U.S. National Oceanic and Atmospheric Administration (NOAA). This analysis shows that: 1)tornados cause the highest number of injuries per year  and also causes the  most fatalities.2)Flood and drought have the most severe econmic damages on properties and crops,respectivly.



#1  Data Loading and  Processing

```{r}
library(ggplot2)
library(R.utils)
library(dplyr)
Sys.setenv(LANG = "en_US.UTF-8")
sessionInfo()
```


 We start the analysis by downloading the data if it is not already in the working directory.
```{r}

if (!file.exists("./StormData")) {dir.create("./StormData")
url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url,"./StormData/dataSet.zip")}


```




#####read in the data. Since the data set 
```{r}
   bunzip2 ("./StormData/dataSet.zip", "./StormData/dataSet.csv")
  stormData<- read.csv ("./StormData/dataSet.csv",sep=",")
```



As the data is very big, We will start by reducing the data set and keep only the columns we need for our analysis.


```{r}
#look in the data
names(StormData)
# we reduce the dimensions of the data set
reducedDf<-stormData[,c(2,8,23,24,25,26,27,28)] 

#remove object "stormData"
rm(stormData)

#view data
head(reducedDf)
```

##### Processing the the BGN_DATE column

```{r}
# remove the "0:00:0" time indication to make the BGN_DATE column easy to handle
reducedDf$BGN_DATE<-gsub(" 0:00:00","",reducedDf$BGN_DATE)

# replace the / by the - 
reducedDf$BGN_DATE<-gsub("/","-",reducedDf$BGN_DATE)

#formate the date
reducedDf$BGN_DATE<-as.Date(reducedDf$BGN_DATE,format = "%m-%d-%Y")

head(reducedDf$BGN_DATE)


```

 
 
 We will replace the "BGN_DATE"" column by a new column with only the year whitout month an day. This step will help to easily reduce the data set by removing observations before 1996.
 
 
```{r}
#select the 4 first characters of the BGN_DATE column
#the 4 first characters correspond to the year
years<-as.numeric(unclass(substr(reducedDf$BGN_DATE, 1, 4)))

#Add the column  "years""
reducedDf<-cbind(reducedDf,years) # add the new column

#remove the "BGN_DATE"column from data set
reducedDf$BGN_DATE <-NULL

#view the data 
dim(reducedDf)

```


##### processing the Columns: FATALITIES & INJURIES & PROPDMG & CROPDMG

We remove all rows where either the pair of fatalities and injuries are both zero, or propertydamage and cropdamage both are zero. 
If a row has PROPDMG = 0, CROPDMG=0, INJURIES+FATALITIES=0 then that observation has no use in this project.  


```{r}
#Subsetting for only the observations where one of those variables is >0 
rmZero<-reducedDf[(reducedDf$FATALITIES > 0 & reducedDf$INJURIES > 0)|reducedDf$PROPDMG  > 0|reducedDf$CROPDMG > 0,]


# uppercasing and cleaned up the PROPDMGEXP and CROPDMGEXP columns so that all the blank values were zero
rmZero$PROPDMGEXP<-toupper(gsub("^$","0",rmZero$PROPDMGEXP))# replace blanks with 0
rmZero$CROPDMGEXP<-toupper(gsub("^$","0",rmZero$CROPDMGEXP))# replace blanks with 0


```


 We create 2 columns with the numeric mulpiers to replace the  CROPDMGEXP and PROPDMGEXP columns. then delete the columns PROPDMGEXP and CROPDMGEXP and calculate the domages caused by the events in dollar. Then Calculage the value of the damages for both crop and propdamgaes
 
 
##### Value of Cropdamages
```{r}
cropExpDf<- data.frame(cropMultiplier=c(1, 1000,  1000000,1000000000), Character = c("0", "K","M","B" ))

cropExpDf # data frame with character & muneric multipliers
MergeCropExp<-merge(rmZero,cropExpDf,by.x="CROPDMGEXP",by.y="Character" )
MergeCropExp$CROPDMGEXP<-NULL # delete the CROPDMGEXP column

# calculate the value of the cropdamage in a colum 
cropDamage<-MergeCropExp$CROPDMG*MergeCropExp$cropMultiplier
MergeCropExp<-cbind(MergeCropExp,cropDamage)
```


##### Value of Propdamages

```{r}
propExpDf<- data.frame(propMultiplier=c(1, 1000,  1000000,1000000000), Characters = c("0", "K","M","B" ))

propExpDf # data frame with character & muneric multipliers
MergedDf<-merge(MergeCropExp,propExpDf,by.x="PROPDMGEXP",by.y="Characters" )
MergedDf$PROPDMGEXP<-NULL # delete the PROPDMGEXP column

# calculate the value of the propdamage in a colum 
propDamage<-MergedDf$PROPDMG*MergedDf$propMultiplier
MergedDf<-cbind(MergedDf,propDamage)


```


##### Processing the column "EVTYPE"
 
We want to  remove white space at the begining of some rows. We have to repeat
 the operation several times since some rows have more than one white spaces at the begining. 
  
```{r}
#  find rows that begin with white space then remove the white space 
MergedDf$EVTYPE<- tolower(MergedDf$EVTYPE) # create a vector of different Event type
for(i in 1:4) {   # repeat the action  times
MergedDf$EVTYPE<-gsub("^ ", "",MergedDf$EVTYPE)
}

```

 there is some duplication of event types under different names. clean up the names of event types:


```{r}
MergedDf$EVTYPE[grep("fire",MergedDf$EVTYPE)]<-"wild fire"     
MergedDf$EVTYPE[grep("forest" ,MergedDf$EVTYPE)]<-"forest fire"     
MergedDf$EVTYPE[grep("erosion" ,MergedDf$EVTYPE)]<-"coastal erosion"     
MergedDf$EVTYPE[grep("glaze" ,MergedDf$EVTYPE)]<-"glaze"     
MergedDf$EVTYPE[grep("torn" ,MergedDf$EVTYPE)]<-"tornado"    
MergedDf$EVTYPE[grep("swells" ,MergedDf$EVTYPE)]<-"high swells"     
MergedDf$EVTYPE[grep("waterspout" ,MergedDf$EVTYPE)]<-"waterspout"     
MergedDf$EVTYPE[grep("microburst" ,MergedDf$EVTYPE)]<-"microburst"     
MergedDf$EVTYPE[grep("lighting" ,MergedDf$EVTYPE)]<-"lightning"     
MergedDf$EVTYPE[grep("freez" ,MergedDf$EVTYPE)]<-"freeze"      
MergedDf$EVTYPE[grep("landslide" ,MergedDf$EVTYPE)]<-"landslide"              
MergedDf$EVTYPE[grep("mudsli",MergedDf$EVTYPE)]<-"mudslide"                 
MergedDf$EVTYPE[grep("cold",MergedDf$EVTYPE)]<-"cold"
MergedDf$EVTYPE[grep("[Ss]now" ,MergedDf$EVTYPE)]<-"snow"
MergedDf$EVTYPE[grep("[Ww]ind" ,MergedDf$EVTYPE)]<-"wind"             
MergedDf$EVTYPE[grep("[Ff]lood",MergedDf$EVTYPE)]<-"flood"
MergedDf$EVTYPE[grep("tstm",MergedDf$EVTYPE)]<-"storm"               
MergedDf$EVTYPE[grep("[Hh]ail" ,MergedDf$EVTYPE)]<-"hail"
MergedDf$EVTYPE[grep("[Hh]eat" ,MergedDf$EVTYPE)]<-"heat"
MergedDf$EVTYPE[grep("[Hh]urricane" ,MergedDf$EVTYPE)]<-"hurricane"
MergedDf$EVTYPE[grep("[Ss]torm",MergedDf$EVTYPE)]<-"storm"
MergedDf$EVTYPE[grep("[Ww]inter" ,MergedDf$EVTYPE)]<-"winter storm"
MergedDf$EVTYPE[grep("[Pp]recip" ,MergedDf$EVTYPE)]<-"precipitation"
MergedDf$EVTYPE[grep("[Cc]urrent" ,MergedDf$EVTYPE)]<-"rip current"
MergedDf$EVTYPE[grep("[Rr]ain" ,MergedDf$EVTYPE)]<-"rain"

dim(MergedDf)

```


the clean up process is over now. the data is tidy. We  can now start the Analysis of the data


# 2  Results



 summarize the  **injuries**, **fatalities**, **crop damages** and **property damages** by event type
 
```{r}

damageDf<- MergedDf %>% group_by(EVTYPE) %>%
      summarize(SumFatalities = sum(FATALITIES), SumInjuries = sum(INJURIES), 
      SumPopDamage = sum(propDamage),SumCropDamage = sum(cropDamage))

dim(damageDf)

````



### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
 
 
```{r}
# view list of the 6 events that cause the most injuries
injDamage<-arrange(damageDf[,c(1,3)],desc(SumInjuries)) %>% head(); 
injDamage

#view list of the 6 event that causes the most fatalies
 fatalDamage<-arrange(damageDf[,c(1,2)],desc(SumFatalities))%>% head()
 fatalDamage


```




### Across the United States, which types of events have the greatest economic consequences?

```{r}
 #view list of  the 6 events that cause the most prop domages
 propDamage<-arrange(damageDf[,c(1,4)],desc(SumPopDamage ))%>% head();
propDamage

# list of the  6 events that cause the most crop damages
 cropDamage<-arrange(damageDf[,c(1,5)],desc(SumCropDamage))%>% head();
cropDamage


```


## Conclusion

Plotting the results
```{r}
 plot1 <- qplot(data=injDamage, x=EVTYPE, y=SumInjuries)

 plot2 <- qplot(data=fatalDamage,x=EVTYPE,y=SumFatalities)
  
 plot3 <- qplot(data=propDamage,x=EVTYPE,y=SumPopDamage)

 plot4 <- qplot(data=cropDamage,x=EVTYPE,y=SumCropDamage)



multiplot <- function(..., plotlist=NULL, cols) {
    require(grid)

    # Make a list from the ... arguments and plotlist
    plots <- c(list(...), plotlist)

    numPlots = length(plots)

    # Make the panel
    plotCols = cols# Number of columns of plots
    
    plotRows = ceiling(numPlots/plotCols) # Number of rows needed, calculated from # of cols

    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(plotRows, plotCols)))
    vplayout <- function(x, y)
        viewport(layout.pos.row = x, layout.pos.col = y)

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
        curRow = ceiling(i/plotCols)
        curCol = (i-1) %% plotCols + 1
        print(plots[[i]], vp = vplayout(curRow, curCol ))
    }

}

multiplot(plot1, plot2, cols=1)
```



This analysis shows that tornado have the highest number of injuries per year  and also causes the  most fatalities.


```{r}
multiplot(plot3, plot4, cols=1)

```

Those events which had great economic consequences are Flood and Drought.
`
